
---
title: "Analysis of world bank population data set"
author: "Uyiosa Igbinedion"
date: '2022-06-13'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```
<br><br>

## Introduction and data set overview 

The data set is from the dplyr package and shows data about world population from the world bank 

The data set has **`r nrow(world_bank_pop)`** rows and **`r ncol(world_bank_pop)`** columns given as: 

1. **country**
    * Three letter country code

1. **indicator**
    * Indicator name: SP.POP.GROW = population growth, SP.POP.TOTL = total population, SP.URB.GROW = urban population growth, SP.URB.TOTL = total urban population

1. **2000-2018**
    * Value for each year


<br><br><br>


## Data transformations

The following transformations would be done on the data to prepare it for analysis


1. Select a subset of the data where `indicator == SP.POP.TOTL | indicator == SP.URB.TOTL`
1. Change the names used in the indicator variable
1. change variable names to title case 
1. Change from wide to long format 
1. Augment data set with additional variables e.g. *Continent*, *Region*, *Income group*, *Actual country name*
1. Drop missing values
1. Create an *others* country category to avoid loosing data 

<br><br><br>


```{r echo=FALSE}
knitr::kable(head(world_bank_pop,5),
             caption = 'First 5 rows of the world bank population dataset')
```



<br><br><br><br>



### Transformation 1 : Select subset of the data

```{r }
#Get only urban and Total population
world_bank_pop_2= filter(world_bank_pop,indicator == "SP.URB.TOTL" | indicator=="SP.POP.TOTL")

```



<br><br><br><br>



### Transformation 2-3 : Change the names used in the indicator variable and title case the variable names 

```{r}
#Change the indicator names to something more understandable
world_bank_pop_2 <- world_bank_pop_2 %>% 
  mutate(indicator= str_replace(indicator,'SP.URB.TOTL','Urban Population'))

world_bank_pop_2 <- world_bank_pop_2 %>% 
  mutate(indicator= str_replace(indicator,'SP.POP.TOTL','Total Population'))

#Change the variable names to title case (capitalize the first letter of the word)
world_bank_pop_2 <- world_bank_pop_2 %>% rename_with(.,str_to_title)

```



<br><br><br><br>



### Transformation 4 : Change from wide to long format

```{r}
#change from wide to long format
world_bank_pop_2 <- world_bank_pop_2 %>% gather(`2000`:`2017`,key="Year",value = "Population")

```



<br><br><br><br>



### Transformation 5 : Augment data set with additional variables

Web scraping of the [World bank](https://wits.worldbank.org/countryprofile/metadata/en/country/all) and [World Population review ](https://worldpopulationreview.com/country-rankings/list-of-countries-by-continent) sites was done in python to augment the data set with additional variables (see untitled2.ipynb)



```{r message=FALSE, warning=FALSE}
#read in additional python-processed data (continent,region,income group,country name)
CountryCode <- read_csv("C:\\Users\\Igbinedion Uyiosa\\Documents\\Uyi\\Data Science\\Datasets\\countryCode.csv")

#joining the two tables into one 
world_bank_pop_3 <-  left_join(CountryCode,world_bank_pop_2, by=c('Country ISO3'='Country'))

```



<br><br><br><br>



### Transformation 6 : Drop missing values

```{r}
#dropping reviewed NA values (1st output --pop4)
world_bank_pop_4 <- world_bank_pop_3 %>% drop_na(Population)

```



<br><br><br><br>



### Transformation 7 : Create an '*others*' country category

```{r message=FALSE, warning=FALSE}
Total_World_population <- as_tibble(world_bank_pop %>% filter(country=='WLD',indicator=='SP.POP.TOTL') %>% select(-country,-indicator) %>% t)
Total_Urban_population <- as_tibble(world_bank_pop %>% filter(country=='WLD',indicator=='SP.URB.TOTL') %>% select(-country,-indicator) %>% t)

Known_World_population <-world_bank_pop_4 %>% group_by(Year) %>% filter(Indicator=='Total Population') %>% summarise(knownPop= sum(Population)) %>% select(knownPop)

Known_Urban_population <-world_bank_pop_4 %>% group_by(Year) %>% filter(Indicator=='Urban Population') %>% summarise(knownPop= sum(Population)) %>% select(knownPop)

totalPopDiff <- Total_World_population - Known_World_population
UrbanPopDiff <- Total_Urban_population - Known_Urban_population

totalPopDiff <- totalPopDiff  %>% mutate(Country_Name = 'Others',Indicator ='Total Population',Year=2000:2017,Country_ISO3='OTH', Region = NA, Continent = NA,Income_Group = NA) %>% rename('Population'='V1') 

UrbanPopDiff <- UrbanPopDiff  %>% mutate(Country_Name = 'Others',Indicator ='Urban Population',Year=2000:2017,Country_ISO3='OTH', Region = NA, Continent = NA,Income_Group = NA) %>%rename('Population'='V1')   

```

<br><br><br><br>

## Visualisations 

Visualizations were done using **Tableau Public**

<br><br><br><br><br><br><br><br>